Parallel-in-time Methods with Machine Learning - Report

### 2020-04-17 report/MSc-report.tex
Update on Overleaf.

\usepackage{epcc}
\usepackage{graphics}
\usepackage{booktabs}
\usepackage{pgfgantt}
\usepackage{url}

@@

Our project focuses on the Parareal algorithm \cite{Parareal} from the family of parallel-in-time methods. Given the necessary computational resources, this algorithm can solve problems faster, in terms of wall clock time, than traditional parallel solvers which can usually only take advantage of limited levels of parallelism tightly bounded by the problem size. This property of the algorithm makes it ideal for problems requiring real-time solutions which is why it is called "Parareal" \cite[p.~3]{staff2003}.

The algorithm works by dividing the time domain up into $P$ slices where $P$ is the number of executing processes or threads. Each one of these $P$ time slices then represents an initial value problem (IVP) that can be solved in parallel. To calculate the initial values, the Parareal algorithm relies on a coarse operator $G$ that is run serially. Once the initial value estimates are computed, the fine operator $F$ is run in parallel across the time slices to solve the individual IVPs and provide correction terms for the subsequent initial values. This is an iterative process where each iteration refines the solution and the total number of iterations necessary is determined by the solution's accuracy requirements.

@@

This brings us to the primary hypothesis of our project which is that a machine learning (ML) model can be trained and used as the coarse operator $G$ in a Parareal framework and it can achieve better performance, in terms of accuracy and execution time, than conventional operators. As there is a substantial body-of-work on using ML to solve differential equations, we begin this report by briefly introducing the most relevant and most notable pieces of research. We then present some of the results of our preliminary investigations into the dissertation project. Following that, we elaborate on our final project proposal, introduce our work plan, present our risk analysis, and outline the contents of the dissertation report. Lastly, we finish with a review of Nestor Sanchez' dissertation from 2018 "Fitting Large-Scale Gaussian Mixtures With Accelerated Gradient Descent" \cite{sanchez2018}.

Parallel-in-time methods, and the Parareal algorithm in particular, have been attracting a lot of interest due to their suitability for long-time simulations and problems requiring fast solutions \cite{gurrala2015}. A lot of this interest has been directed to the analysis of the algorithm itself \cite{staff2003} \cite{gander2007}. An important property of the Parareal algorithm is that it converges to the serial solution as computed by $F$ in a maximum of $P$ iterations \cite[p.~6-7]{staff2003}. However, for optimal speedup, the algorithm needs to converge in much fewer iterations than $P$. In fact, if the computational cost of integrating over the entire domain using the operator $G$ matches the cost of integrating over a single time slice using $F$, the algorithm has to converge in fewer than $P / 2$ iterations to achieve any speedup \cite[p.~7]{staff2003}.

@@

In contrast, others have attempted to tackle the issue directly by experimenting with different coarse operators. Generally, a simple Runge-Kutta method \cite{runge1895} such as Euler's method is used as $G$ and a higher order method, potentially with finer temporal discretisation, such as the Runge-Kutta 4th order (RK4) method is used as $F$ \cite[p.~1822]{gurrala2015} \cite[p.~2]{duan2016}. However, it has been shown that there is more to consider than just the integration method and its step size. In some cases, it is possible to improve the speed of the coarse operator without a significant detriment to accuracy by simplifying the differential equation (only for $G$) and thus reducing the complexity of the model it operates on \cite{duan2016}. Consequently, selecting and fine tuning the coarse operator is a challenging task that boils down to finding or building a differential equation solver with optimal speed and accuracy characteristics for the problem at hand.

Given the success of ML in a wide range of computational fields, it is unsurprising that their application to the solution of differential equations is a popular field of research. Due to their prowess as universal function approximators \cite{hornik1991}, most efforts are targeted towards using artificial neural networks (ANN) to approximate the solutions of differential equations \cite{lagaris1998} \cite{lagaris2000} \cite{han2018}. While the majority of these works take advantage of the differentiability of neural networks and use training processes tailored explicitly to them, it has been shown that it is possible to use ML models to approximate differential equations in a strictly supervised, black-box setting as well \cite{regazzoni2019}.

Perhaps the most successful approach so far to solve complex differential equations with ML is based on so called physics informed neural networks (PINN) \cite{raissi2017} \cite{tartakovsky2018} \cite{sirignano2018}. PINNs are architecturally standard neural networks with the values of the differential equation's independent variables ($t$ and $x$) as their input and the value of the solution at the point defined by the independent variables as their output. The most important feature of PINNs is their specialised loss function that penalises violations of the initial conditions, the boundary conditions, and the differential equation itself in the form of L2 terms. This loss function allows the network to be trained through gradient descent in a largely unsupervised fashion on collocation points sampled from the differential equation's spatiotemporal domain according to a multivariate distribution (with an additional set of points sampled from the boundaries of the spatial domain according to a different distribution). While PINNs are highly efficient for certain types of high dimensional partial differential equations (PDE), for many problems they are still slower than conventional solvers such as the finite element method (FEM) \cite{lu2019}.

@@

The first one of these milestones has been the implementation of a basic 1D diffusion simulation in C. This simulation uses simple Dirichlet boundary conditions and Gaussian initial values. The spatial discretisation is done via the finite difference method and the integrator used is Euler's method.

The second, somewhat more ambitious milestone has been the implementation of a simple shared-memory Parareal framework in C using POSIX threads to solve an ordinary differential equation (ODE) describing the population growth of rabbits. This framework employs Euler's method as $G$, RK4 as $F$, and can only handle matching coarse and fine operator step sizes. Although the algorithm runs for a predefined number of corrective iterations, the implementation demonstrates a significant speedup at reasonable levels of accuracy even when running on only 4 threads. Moreover, it showcases how the parallel solution converges to the serial fine solution as the number of iterations approaches $P$.

Finally, our third and most substantial milestone has been the development of a distributed Parareal framework in Python using mpi4py \cite{dalcin2005}. This framework is fairly flexible in that it can handle any operator and any ODE that conform to the respective interface definitions. The algorithm also supports early stopping based on the absolute value of the largest update to the solution at any time step. In addition, the framework provides a number of Runge-Kutta method implementations and a conventional operator that takes an integrator instance and a step size value as its parameters. Furthermore, the framework also includes an ML accelerated operator implementation that is parameterised by a regression model, a target operator, and a step size. The ML operator trains the model on every new differential equation by generating a set of input-output pairs through a stochastic process using the target operator and fitting the regression model to the generated data. Using this approach, we have successfully trained and utilised an ML operator backed by a linear regression model as the coarse operator $G$ in our Parareal framework to solve a simple ODE.

Through these milestones, we have improved our understanding of the Parareal algorithm and gained a bit of experience in applied numerical analysis. Moreover, we have also managed to confirm the project's feasibility on a highly simplified problem by showing that an ML accelerated coarse operator can outperform a conventional one. Finally, we have created a strong foundation and starting point for our dissertation project.

@@

Our final project proposal is based closely on the key steps outlined in the original proposal and it introduces only minor changes. The focus remains implementing a Parareal framework using an ML model as the coarse operator and solving differential equations of increasing complexity. Similarly, the primary metrics are still the wall time, computational cost, and accuracy of the coarse operator and the framework as a whole. Consequently, we formulate our research questions as

\begin{enumerate}
    \item how does an ML accelerated coarse operator compare to traditional coarse operators in terms of wall time and accuracy
    \item is an ML accelerated Parareal framework faster than a conventional Parareal framework given the same accuracy requirements
    \item is an ML accelerated Parareal framework faster than other conventional parallel solvers given the same accuracy requirements and using the same number of parallel processes
\end{enumerate}

@@

In addition, we also intend to try sequence models such as recurrent neural networks (RNN) to capture temporal patterns in the differential equations' solutions. We hypothesize that this increases the robustness of the learned models and aids the approximation of periodic solutions. However, this requires the capturing of inputs from possibly multiple previous time steps. Therefore, we propose a more general training and evaluation scheme where the input is a sequence. The training of the non-recurrent models would be, thus, merely a special case of this scheme where the input sequence contains only one element. This allows for the implementation of a single ML accelerated operator that is agnostic of the model.

@@

We propose implementing our operators in Python using sklearn \cite{pedregosa2011} for linear regression and the decision tree based models and either Tensorflow \cite{mart2015} and Keras \cite{chollet2015} or PyTorch \cite{paszke2019} for the neural network models. As for our benchmarks, we plan to compare the execution time and accuracy of the ML accelerated operators to conventional solvers provided by scipy \cite{scipy2020}, FEniCS \cite{fenics2015}, and FiPy \cite{fipy2009} (depending on the type of the differential equation). Additionally, we are also interested in the training times of the ML operators using different models. For differential equations without an analytic solution, we base our accuracy comparisons on the solutions of higher order, small step-size solvers.

@@

For the comparison between the ML accelerated and the normal versions of our Parareal framework, we are interested in the wall time of the solution using the same number of processes and the same accuracy requirements. To ensure our analysis is comprehensive enough, we propose performing comparisons across different combinations of process counts and accuracy levels. We intend to time the ML accelerated version both including and excluding the training of the models. Furthermore, we also aim to consider the amortized cost of model training through transfer learning. We plan to do so by quantifying how well a model trained on one IVP performs on a different IVP expressed by the same differential equation. We propose using the same approach to answer the third research question as well where we compare our ML accelerated Parareal framework to conventional parallel ODE and PDE solvers provided by FEniCS.

Finally, as for the differential equations to benchmark our operators and framework on, we set forth three different problems of increasing complexity. The first one is the Lotka-Volterra model \cite{lotka1926} \cite{volterra1927} which is a system of two ODEs expressing the dynamics of the populations of predators and their preys. This requires the extension of our initial framework to handle systems of differential equations which is the same as handling differential equations with vector valued solutions. The second differential equation we propose solving is the two dimensional heat equation. This requires the extension of the framework to handle PDEs, that is differential equations of multiple independent variables. Lastly, the third problem we consider is the Black-Scholes equation \cite{black1973} which is a high dimensional PDE that can be computationally challenging to solve using traditional mesh-based approaches.

@@

In this section, we outline our work plan for the dissertation period beginning on 25th May and ending on 21st August. The units of work listed below are based on our final project proposal.

\begin{enumerate}
    \item Implementation
        \begin{enumerate}
        	\item Testing
        	\item Systems of ODEs
        	\item PDEs
        	\item ML operators
        	\item Optimisation
        \end{enumerate}
    \item Benchmarking
        \begin{enumerate}
        	\item Only ML operators
        	\item Framework with and without ML
        	\item Against another parallel solver
        \end{enumerate}
    \item Writing
        \begin{enumerate}
            \item Engineering log
            \item Phase 1
            \item Phase 2
        	\item Proof reading
        \end{enumerate}
\end{enumerate}

The three main tasks of the dissertation we identify are the implementation of the software framework, the benchmarking of its performance, and the write up of the dissertation report. Each one of these high level tasks are divided up into smaller sub-tasks. For the implementation task, these sub-tasks are software testing, the extension of the Parareal framework to handle systems of ODEs and individual PDEs, the implementation of the two final ML accelerated operators discussed in the project proposal (one for PINNs and one for all the other models), and code optimisation. As for the benchmarking task, the sub-tasks are the comparison of the ML accelerated operator to conventional operators, the comparison of the Parareal framework using an ML accelerated coarse operator to the same framework using a conventional coarse operator, and the comparison of the ML accelerated Parareal framework to other parallel solvers. Finally, the last main task, the writing of the dissertation consists of four smaller units of work. These are the engineering log to keep during the initial development period, the first phase for writing up the problem formulation and the design and implementation sections of the dissertation, the second phase for writing up the rest of the dissertation such as the literature review and the benchmark results, and lastly, the proof reading of the report.

\begin{figure}[!htb]
\begin{center}
\ganttset{calendar week text={W\currentweek}}
\begin{ganttchart}[
        hgrid,
        vgrid,
        x unit=1mm,
        y unit chart=5mm,
        time slot format=isodate,
        time slot unit=day]{2020-05-25}{2020-08-23}
    \gantttitlecalendar{month=name, week=1} \\
    \ganttgroup{Implementation}{2020-05-25}{2020-07-18} \\
    \ganttbar[bar/.style={fill=red!50}]{Testing}{2020-05-25}{2020-06-28} \\
    \ganttbar[bar/.style={fill=red!50}]{Systems of ODEs}{2020-05-25}{2020-06-07} \\
    \ganttbar[bar/.style={fill=red!50}]{PDEs}{2020-05-31}{2020-06-14} \\
    \ganttbar[bar/.style={fill=red!50}]{ML operators}{2020-06-15}{2020-06-28} \\
    \ganttbar[bar/.style={fill=red!50}]{Optimisation}{2020-06-29}{2020-07-18} \\
    \ganttgroup{Benchmarking}{2020-07-04}{2020-07-17} \\
    \ganttbar[bar/.style={fill=green!50}]{Only ML operators}{2020-07-04}{2020-07-10} \\
    \ganttbar[bar/.style={fill=green!50}]{Framework with and without ML}{2020-07-08}{2020-07-13} \\
    \ganttbar[bar/.style={fill=green!50}]{Against another parallel solver}{2020-07-10}{2020-07-17} \\
    \ganttgroup{Writing}{2020-05-25}{2020-08-21} \\
    \ganttbar[bar/.style={fill=blue!50}]{Engineering log}{2020-05-25}{2020-06-28} \\
    \ganttbar[bar/.style={fill=blue!50}]{Phase 1}{2020-06-29}{2020-08-02} \\
    \ganttbar[bar/.style={fill=blue!50}]{Phase 2}{2020-08-03}{2020-08-20} \\
    \ganttbar[bar/.style={fill=blue!50}]{Proof read}{2020-08-16}{2020-08-21}
\end{ganttchart}
\end{center}
\caption{Gantt chart of the dissertation work plan}
\label{fig:gantt}
\end{figure}

Figure \ref{fig:gantt} contains a Gantt chart displaying the timeline of the dissertation project and the different tasks it encompasses. Although the times are just rough and somewhat optimistic estimates, the chart sufficiently visualises the overlaps between certain tasks. For example, it clearly captures the non-linear nature of optimisation and benchmarking as they feed into each other.

@@

We identify the main risks of the dissertation project as

\begin{itemize}
    \item Failure to implement a single framework to handle both ODEs and PDEs
    \item Failure to fit a regression model to complex differential equations
    \item Failure to utilise transfer learning
    \item Unavailability of computing resources
    \item Disruptions due to Covid-19
\end{itemize}

>>>>

\usepackage{epcc}
\usepackage{graphics}
\usepackage{booktabs}
\usepackage{pgfgantt}
\usepackage{url}
\usepackage{hyperref}
\usepackage[thinc]{esdiff}
\usepackage{tabularx}

\newcolumntype{Y}{>{\centering\arraybackslash}X}

@@

Our project focuses on the Parareal algorithm \cite{staff2003} from the family of parallel-in-time methods. Given the necessary computational resources, this algorithm can solve problems faster, in terms of wall clock time, than traditional parallel solvers which can usually only take advantage of limited levels of parallelism tightly bounded by the problem size. This property of the algorithm makes it ideal for problems requiring real-time solutions which is why it is called "Parareal" \cite[p.~3]{staff2003}.

The Parareal algorithm solves time-dependent differential equations by splitting them up along the time domain and assigning a sub-problem to each processor. It first computes the initial values for each sub-problem serially using a coarse operator. It then solves each of these problems based on the initial values in parallel using a fine operator and calculates a correction to the initial values. This iterative process goes on until a sufficient level of accuracy is reached. As the coarse operator is executed serially at every iteration, it has to be fast but it also needs to be accurate enough for rapid convergence.

This brings us to the primary hypothesis of our project which is that a machine learning (ML) model can be trained and used as the coarse operator in a Parareal framework and it can achieve better performance, in terms of accuracy and execution time, than conventional operators. As there is a substantial body-of-work on using ML to solve differential equations, we begin this report by briefly introducing the most relevant and most notable pieces of research. We then present some of the results of our preliminary investigations into the dissertation project. Following that, we elaborate on our final project proposal, introduce our work plan, present our risk analysis, and outline the contents of the dissertation report. Lastly, we finish with a review of Nestor Sanchez' dissertation from 2018 "Fitting Large-Scale Gaussian Mixtures With Accelerated Gradient Descent" \cite{sanchez2018}.

The Parareal algorithm works by dividing the time domain up into $P$ slices where $P$ is the number of executing processes or threads. Each one of these $P$ time slices then represents an initial value problem (IVP) that can be solved in parallel. To calculate the initial values, the Parareal algorithm relies on a coarse operator, $G$, that is run serially. Once the initial value estimates are computed, the fine operator, $F$, is run in parallel across the time slices to solve the individual IVPs and provide correction terms for the subsequent initial values. This is an iterative process where each iteration refines the solution and the total number of iterations necessary is determined by the solution's accuracy requirements.

@@

The Parareal algorithm has been attracting a lot of interest due to its suitability for long-time simulations and problems requiring fast solutions \cite{gurrala2015}. A lot of this interest has been directed to the analysis of the algorithm itself \cite{staff2003} \cite{gander2007}. An important property of the algorithm is that it converges to the serial solution as computed by $F$ in a maximum of $P$ iterations \cite[p.~6-7]{staff2003}. However, for optimal speedup, it needs to converge in much fewer iterations than $P$. In fact, if the computational cost of integrating over the entire domain using the operator $G$ matches the cost of integrating over a single time slice using $F$, the algorithm has to converge in fewer than $P / 2$ iterations to achieve any speedup \cite[p.~7]{staff2003}.

@@

In contrast, others have attempted to tackle the issue directly by experimenting with different coarse operators. Generally, a simple Runge-Kutta method such as the forward Euler method is used as $G$ and a higher order method, potentially with finer temporal discretisation, such as the Runge-Kutta 4th order (RK4) method is used as $F$ \cite[p.~1822]{gurrala2015} \cite[p.~2]{duan2016}. However, it has been shown that there is more to consider than just the integration method and its step size. In some cases, it is possible to improve the speed of the coarse operator without a significant detriment to accuracy by simplifying the differential equation (only for $G$) and thus reducing the complexity of the model it operates on \cite{duan2016}. Consequently, selecting and fine tuning the coarse operator is a challenging task that boils down to finding or building a differential equation solver with optimal speed and accuracy characteristics for the problem at hand.

Given the success of ML in a wide range of computational fields, it is unsurprising that their application to the solution of differential equations is a popular field of research. The most attractive prospect of using ML to this end is that a trained model can directly represent the solution without the need for any discretisation (which can be intractable for high dimensional problems). Due to their prowess as universal function approximators \cite{hornik1991}, most efforts are targeted towards using artificial neural networks (ANN) to approximate the solutions of differential equations \cite{lagaris1998} \cite{lagaris2000} \cite{han2018}. This is also partly attributable to the fact that ANNs are differentiable which allows for more efficient training approaches. While the majority of recent works take advantage of the differentiability of neural networks and use training processes tailored explicitly to them, it has been shown that it is possible to use ML models to approximate the solutions of differential equations in a strictly supervised, black-box setting as well \cite{regazzoni2019}.

Perhaps the most successful approach so far to solve complex differential equations with ML is based on physics informed neural networks (PINN) \cite{raissi2017} \cite{tartakovsky2018} \cite{sirignano2018}. PINNs approximate the function that solves the differential equation, therefore their inputs are the values of the differential equation's independent variables ($t$ and $x$) and their output is the value of the solution at the point defined by the inputs. In terms of architecture, PINNs are flexible and can be implemented by simple ANNs, convolutional neural networks, and even recurrent neural networks (RNN). The most important feature of PINNs is their specialised loss function that penalises violations of the initial conditions, the boundary conditions, and the differential equation itself in the form of mean squared error terms. By minimising this loss function, for example through gradient descent, the network can be trained in a largely unsupervised fashion on collocation points sampled from the differential equation's spatiotemporal domain according to a multivariate distribution (with an additional set of points sampled from the boundaries of the domain according to a different distribution). Although training a PINN can be very costly, especially on differential equations with a large domain, a trained model may perform more than an order of magnitude faster than conventional solvers at comparable levels of accuracy \cite{misyris2019}. However, while PINNs are fully mesh-free and very efficient for high dimensional partial differential equations (PDE), for some problems they are still slower than conventional solvers such as the finite element method \cite{lu2019}.

@@

\begin{equation}
    \diffp{u}{t} = d \diffp{u}{{x^2}}
\label{equation:1d_diff}
\end{equation}

The first one of these milestones has been the implementation of a basic one dimensional diffusion simulation in C defined by the model in Equation \ref{equation:1d_diff} where $u$ is the unknown function, $x$ is the spatial coordinate, $t$ is time, and $d$ is the diffusion coefficient. This simulation uses simple Dirichlet boundary conditions and Gaussian initial values. The spatial discretisation is done via the finite difference method and the integrator used is the forward Euler method. The implementation can be found at \href{https://git.ecdf.ed.ac.uk/msc-19-20/s1984842/blob/master/code/c/src/main.c}{\texttt{code/c/src/main.c}} in the project repository.

\begin{equation}
    \diff{R}{t} = \alpha R
\label{equation:rabbit_pop}
\end{equation}

The second, somewhat more ambitious milestone has been the implementation of a simple shared-memory Parareal framework in C using POSIX threads to solve an ordinary differential equation (ODE) describing the population growth of rabbits. This ODE is expressed by Equation \ref{equation:rabbit_pop} where $R$ is the number of rabbits and $\alpha$ is the population growth rate. The framework employs the forward Euler method as $G$ and RK4 as $F$ but it can only handle matching coarse and fine operator step sizes. Although the algorithm runs for a predefined number of corrective iterations, the implementation demonstrates a significant speedup at reasonable levels of accuracy even when running on only 4 threads. Moreover, it showcases how the parallel solution converges to the serial fine solution as the number of iterations approaches $P$. This implementation can be found in the same file as the diffusion simulation.

Finally, our third and most substantial milestone has been the development of a distributed Parareal framework in Python using mpi4py \cite{dalcin2005}. This framework is fairly flexible in that it can handle any operator and any ODE that conform to the respective interface definitions. The algorithm also supports early stopping based on the absolute value of the largest update to the solution at any time step. In addition, the framework provides a number of Runge-Kutta method implementations and a conventional operator that takes an integrator instance and a step size value as its parameters. Furthermore, the framework also includes an ML accelerated operator implementation that is parameterised by a regression model, a target operator, and a step size. The ML operator trains the model on every new differential equation by generating a set of input-output pairs through a stochastic process using the target operator and fitting the regression model to the generated data. The data generation involves repeatedly integrating over the differential equation by the target operator and adding a $0$-mean Gaussian noise to the estimate at every time step to use as the initial value for the next step and thus perturbate the trajectory of the target operator's solution. Using this approach, we have successfully trained and utilised an ML operator backed by a linear regression model as the coarse operator $G$ in our Parareal framework to solve a simple ODE. The implementation of this framework and our experiment can be found in the \href{https://git.ecdf.ed.ac.uk/msc-19-20/s1984842/blob/master/code/python/}{\texttt{code/python}} directory of the project repository

\begin{table}[!htb]
\begin{center}
\begin{tabularx}{\textwidth}{Y|Y|Y|}
\multicolumn{1}{c}{} & \multicolumn{2}{c}{Operator}\\
\cmidrule(l){2-3}
& Explicit midpoint & Linear regression\\
\midrule
Run time (s) & 0.233 & \textbf{0.017}\\
\bottomrule
\end{tabularx}
\end{center}
\caption{The average execution times of different coarse operators on the rabbit population problem}
\label{tab:coarse_times}
\end{table}

Table \ref{tab:coarse_times} shows the average execution times, based on $10$ runs, of two different coarse operators as they integrate over the entire time domain of the rabbit population problem defined by Equation \ref{equation:rabbit_pop} using the values $R_0 = 1000$, $\alpha = 0.0001$, and $t \in [0, 50000]$. The first operator is a conventional one using the explicit midpoint integration method and a step size of $0.25$ while the second operator is an ML accelerated one with a step size of $100$ backed by an sklearn \cite{pedregosa2011} linear regression model trained on data labelled by the first operator. As seen in the table, the ML accelerated operator solves the differential equation almost $14$ times faster with an average deviation of less than $10^{-6}$ from the conventional operator's solution. This performance difference is due to the fact that the ML operator's step size is $400$ times greater and therefore the solution is evaluated $400$ times fewer while still maintaining the same level of accuracy.

\begin{table}[!htb]
\begin{center}
\begin{tabularx}{\textwidth}{Y|Y|Y|Y|Y|}
\multicolumn{1}{c}{} & \multicolumn{4}{c}{Solver}\\
\cmidrule(l){2-5}
& Serial & Parareal & ML Parareal & ML Parareal with training\\
\midrule
Run time (s) & 8.387 & 6.082 & \textbf{5.080} & 8.020\\
\bottomrule
\end{tabularx}
\end{center}
\caption{The average execution times of different solvers on the rabbit population problem}
\label{tab:parareal_times}
\end{table}

Table \ref{tab:parareal_times} shows the average execution times, again across $10$ runs, of different solvers on the same rabbit population problem. The first, serial solver is a conventional operator using the RK4 integration method and a step size of $0.01$. The others are all Parareal solvers running on $4$ processes using the first solver as their fine operator and $0.001$ as their update threshold (which means that the algorithm finishes once the absolute value of the largest update to the solution at any point of the fine operator's temporal discretisation is below this threshold). The first Parareal solver uses the explicit midpoint operator from Table \ref{tab:coarse_times} as its coarse operator while the second and third Parareal solvers use the linear regression operator. Both the conventional and the ML accelerated Parareal solvers converge after just one corrective iteration and thus are able to achieve a significant speedup compared to the serial solver while producing a solution identical to the serial one to $6$ and $5$ significant figures respectively. While the ML accelerated operator is slightly less accurate, it still converges within the same number of iterations given our accuracy threshold and it actually solves the differential equation a whole second faster. In contrast, when taking the model training into account, the ML accelerated Parareal solver is barely  faster than the serial solver. However, the training time could potentially be significantly reduced for the given accuracy requirements by decreasing the volume of training data generated.

Through these investigations and experiments presented above, we have improved our understanding of the Parareal algorithm and gained a little experience in applied numerical analysis. Moreover, we have also managed to confirm the project's feasibility on a highly simplified problem by showing that an ML accelerated coarse operator can outperform a conventional one. Finally, we have created a strong foundation and starting point for our dissertation project.

@@

The main focus of our project is implementing a Parareal framework using an ML model as the coarse operator and solving differential equations of increasing complexity. The primary metrics we are interested in are the wall time, computational cost, and accuracy of the coarse operator and the framework as a whole. Consequently, we formulate our research questions as

\begin{enumerate}
    \item How do ML accelerated coarse operators compare to traditional coarse operators in terms of accuracy, execution time, and training time?
    \item Is an ML accelerated Parareal framework faster, both excluding and including model training time, than a conventional Parareal framework given the same accuracy requirements?
    \item Is an ML accelerated Parareal framework faster, both excluding and including model training time, than other conventional parallel solvers given the same accuracy requirements and using the same number of parallel processes?
\end{enumerate}

@@

In addition, we also intend to try sequence models such as RNNs to capture temporal patterns in the differential equations' solutions. We hypothesize that this increases the robustness of the learned models and aids the approximation of periodic solutions. However, this requires the capturing of inputs from possibly multiple previous time steps. Therefore, we propose a more general training and evaluation scheme where the input is a sequence. The training of the non-recurrent models is, thus, merely a special case of this scheme where the input sequence contains only one element. This allows for the implementation of a single ML accelerated operator that is agnostic of the model.

@@

While we are optimistic about machine learning accelerated operators with a greater step size outperforming conventional operators in terms of execution time at similar levels of accuracy, we are conscious of the overhead introduced by training data generation and model fitting. Therefore, we propose measuring the average training time for all the models tested. Moreover, due to the inherently stochastic nature of the training processes, we propose basing all our results on the means and standard deviations of measurements across a high number of runs and using fixed random seeds to facilitate reproducibility. Even though the combined training time and execution time of a machine learning operator is more than likely to significantly exceed the execution time of a conventional operator, this may be mitigated in a Parareal framework where a model is trained only once but used for inference potentially several times. Therefore, by keeping track of the training times of the models, we can establish a more complete picture of their performance potential as coarse operators in a Parareal setting.

We propose implementing our operators in Python using sklearn for linear regression and the decision tree based models and either Tensorflow \cite{mart2015} and Keras \cite{chollet2015} or PyTorch \cite{paszke2019} for the neural network models. As for our benchmarks, we plan to compare the execution time and accuracy of the ML accelerated operators to conventional solvers provided by scipy \cite{scipy2020}, FEniCS \cite{fenics2015}, and FiPy \cite{fipy2009} (depending on the type of the differential equation).

@@

For the comparison between the ML accelerated and the normal versions of our Parareal framework, we are interested in the wall time of the solution using the same number of processes and the same accuracy requirements. To ensure our analysis is comprehensive enough, we propose performing comparisons across different combinations of process counts and accuracy levels. We intend to time the ML accelerated version both including and excluding the training of the models. Furthermore, as a stretch goal, we also aim to consider the amortized cost of model training through transfer learning. We plan to do so by quantifying how well a model trained on one IVP performs on a different IVP expressed by the same differential equation. We propose using the same approach to answer the third research question as well where we compare our ML accelerated Parareal framework to conventional parallel ODE and PDE solvers provided by FEniCS. For differential equations without an analytic solution, we plan to base our accuracy comparisons on the solutions of higher-order, smaller-step size solvers.

Finally, as for the differential equations to benchmark our operators and framework on, we set forth three different problems of increasing complexity. The first one is the Lotka-Volterra model \cite{lotka1926} which is a system of two ODEs expressing the dynamics of the populations of predators and their preys. This requires the extension of our initial framework to handle systems of differential equations which is the same as handling differential equations with vector valued solutions. The second differential equation we propose solving is the two dimensional heat equation. This requires the extension of the framework to handle PDEs, that is differential equations of multiple independent variables. Lastly, the third problem we consider is the Black-Scholes equation \cite{black1973} which is a high dimensional PDE that can be computationally challenging to solve using traditional mesh-based approaches.

@@

In this section, we outline our work plan for the dissertation period beginning on 25th May and ending on 21st August. The three main tasks of the dissertation we identify, based on our final project proposal, are the implementation of the software framework, the benchmarking of its performance, and the write up of the dissertation report. Each one of these high level tasks are divided up into smaller sub-tasks.

\begin{enumerate}
    \item Implementation
        \begin{enumerate}
        	\item Testing
        	\item Systems of ODEs
        	\item PDEs
        	\item ML operators
        	\item Performance optimisation
        \end{enumerate}
    \item Benchmarking
        \begin{enumerate}
        	\item Only ML operators
        	\item Framework with and without ML
        	\item Against another parallel solver
        \end{enumerate}
    \item Writing
        \begin{enumerate}
            \item Engineering log
            \item Phase 1
            \item Phase 2
        	\item Proof reading
        \end{enumerate}
\end{enumerate}

For the implementation task, these sub-tasks are software testing, the extension of the Parareal framework to handle systems of ODEs and individual PDEs, the implementation of the two final ML accelerated operators discussed in the project proposal (one for PINNs and one for all the other models), and code optimisation along with hyperparameter tuning. As for the benchmarking task, the sub-tasks are the comparison of the ML accelerated operator to conventional operators, the comparison of the Parareal framework using an ML accelerated coarse operator to the same framework using a conventional coarse operator, and the comparison of the ML accelerated Parareal framework to other parallel solvers. Finally, the last main task, the writing of the dissertation consists of four smaller units of work. These are the engineering log to keep during the initial development period, the first phase for writing up the problem formulation and the design and implementation sections of the dissertation, the second phase for writing up the rest of the dissertation such as the literature review and the benchmark results, and lastly, the proof reading of the report.

\begin{figure}[!htb]
\begin{center}
\ganttset{calendar week text={W\currentweek}}
\begin{ganttchart}[
        hgrid,
        vgrid,
        x unit=1mm,
        y unit chart=5mm,
        time slot format=isodate,
        time slot unit=day]{2020-05-25}{2020-08-23}
    \gantttitlecalendar{month=name, week=1} \\
    \ganttgroup{Implementation}{2020-05-25}{2020-07-18} \\
    \ganttbar[bar/.style={fill=red!50}]{Testing}{2020-05-25}{2020-06-28} \\
    \ganttbar[bar/.style={fill=red!50}]{Systems of ODEs}{2020-05-25}{2020-06-07} \\
    \ganttbar[bar/.style={fill=red!50}]{PDEs}{2020-05-31}{2020-06-14} \\
    \ganttbar[bar/.style={fill=red!50}]{ML operators}{2020-06-15}{2020-06-28} \\
    \ganttbar[bar/.style={fill=red!50}]{Performance otimisation}{2020-06-29}{2020-07-18} \\
    \ganttgroup{Benchmarking}{2020-07-04}{2020-07-17} \\
    \ganttbar[bar/.style={fill=green!50}]{Only ML operators}{2020-07-04}{2020-07-10} \\
    \ganttbar[bar/.style={fill=green!50}]{Framework with and without ML}{2020-07-08}{2020-07-13} \\
    \ganttbar[bar/.style={fill=green!50}]{Against another parallel solver}{2020-07-10}{2020-07-17} \\
    \ganttgroup{Writing}{2020-05-25}{2020-08-21} \\
    \ganttbar[bar/.style={fill=blue!50}]{Engineering log}{2020-05-25}{2020-06-28} \\
    \ganttbar[bar/.style={fill=blue!50}]{Phase 1}{2020-06-29}{2020-08-02} \\
    \ganttbar[bar/.style={fill=blue!50}]{Phase 2}{2020-08-03}{2020-08-20} \\
    \ganttbar[bar/.style={fill=blue!50}]{Proof reading}{2020-08-16}{2020-08-21}
\end{ganttchart}
\end{center}
\caption{Gantt chart of the dissertation work plan}
\label{fig:gantt}
\end{figure}

Figure \ref{fig:gantt} contains a Gantt chart displaying the timeline of the dissertation project and the different tasks it encompasses. Although the times are just rough and somewhat optimistic estimates, the chart sufficiently visualises the overlaps between certain tasks. For example, it clearly captures the non-linear nature of optimisation and benchmarking as they feed back into each other.

As for hardware requirements, our plan is to use the author's personal computer for most of the implementation phase. This computer has $6$ physical cores which is sufficient for development and testing purposes. However for optimisation and benchmarking, we plan to use high performance computing services. Due to the generally low number of processors needed for our benchmarks, we estimate our required budget to be about $2000$ to $3000$ CPU core hours on Cirrus. Additionally, if the new Cirrus GPUs are installed by late June, we also plan to take advantage of them for our neural network based models. For this, our required budget estimate is $50$ GPU hours.

@@

We identify the following as the main risks of the dissertation project.

\begin{itemize}
    \item Failure to implement a single framework to handle both ODEs and PDEs in time
    \item Some of the model training approaches may not work on complex problems
    \item Erratic model behaviour due to the stochastic nature of the training processes
    \item Transfer learning may not work
    \item Unavailability of computing resources
    \item Disruptions due to Covid-19
\end{itemize}

The first risk we identify is the author's failure to implement a Parareal framework that can handle ODEs, systems of ODEs, and PDEs in time for our benchmarks that need to be performed early enough to leave sufficient time for writing up the results. Failure to do this would likely restrict our benchmarks to ODEs only which are not representative of more complex real life problems. However, this risk is mitigated by the fact that we already have a strong base in the form of our Parareal framework implementation that can handle any ODE. Moreover, we have a good idea of how to extend this framework to enable systems of ODEs and PDEs as well.

The second major risk is that some of the models or training processes we intend to try may simply not work. While this is a high probability risk, it is rather low impact due to the fact that we plan to try multiple approaches, some of which have already been shown to work on several problems (such as PINNs). Additionally, we have already succeeded in training an ML accelerated operator that can outperform some conventional solvers on a very simple ODE.

The third risk is erratic model behaviour due to the stochastic nature of training. As both our proposed supervised training approach and the unsupervised training method of PINNs have elements of randomness (and the weights of neural networks are initialised randomly as well), the models may behave fairly differently between runs. To address this, we plan to base our results on high numbers of runs. For the sake of reproducibility, we also intend to use fixed random seeds in our benchmarks.

The next main risk we identify is that transfer learning may not work. This means that a model trained on a smaller subset of a differential equation's domain may not be able to infer the solution well enough outside of that subset. Alternatively, a model trained on a differential equation using certain initial value and boundary conditions may not do well on the same equation with different conditions. This is another high probability risk, but utilising transfer learning is only a secondary goal of our project and we believe that our results would be valuable even without it.
